---
title: "Practical"
subtitle: "Measuring Biodiversity and inferring Ecosystem Processes"  
author: 
  - "Jasper Slingsby, BIO3018F"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current% / %total%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)

library(RefManageR)
BibOptions(check.entries = FALSE,
           bib.style = "authoryear",
           cite.style = "alphabetic",
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)
myBib <- ReadBib("bib/2_species.bib", check = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)

# style_duo_accent(
#   primary_color = "#1381B0",
#   secondary_color = "#FF961C",
#   inverse_header_color = "#FFFFFF"
# )

style_mono_light(base_color = "#23395b")

#https://mycolor.space/?hex=%2323395B&sub=1 
#"Generic gradient" - #23395B #006287 #008E9D #00B897 #89DD81 #F9F871
#"Matching gradient" (reverse) - #23395B #494E77 #716292 #9C77AA #C88DBF #F5A3D0



library(knitr)
library(kableExtra)
```


```{r xaringan-tile-view, echo=FALSE}
# xaringanExtra::use_tile_view()
```

class: center, middle

## Scenario

We are on a newly discovered island. In order to manage the island's biodiversity, we need to do a **quantitative** survey of the plant communities so as to describe their biodiversity and infer the key ecological processes that drive them.

---

## This can be broken down into the following questions:

- How do communities differ in terms of their $\alpha$ and $\beta$ species and functional diversity and is there reason to group or differentiate some relative to others?

- What are the dominant ecological processes in each community and how could you infer them?

- Is there an association between the variation in environmental conditions and/or ecological processes and the observed variation in biodiversity (species and functional $\alpha$ and $\beta$ diversity) that may suggest a causal link?


---

## For each question: 

Please propose hypotheses and predictions that allow you to test each hypothesis (hint: see Claire's lecture).

Also consider and discuss what factors, if any, could confound your ability to answer the question. Have you been able to account for them? If not, what would you need to be able to do so?

---

class: center

## The De Hoop landscape includes a wide range of environmental conditions and vegetation communities

```{r echo = F, fig.align = 'center', out.height = '60%', warning=F, messages=F}

library(terra, quiet = T)
library(sf, quiet = T)
library(tidyverse, quiet = T)
library(readxl, quiet = T)
library(mapview, quiet = T)
library(vegan, quiet = T)
library(GGally, quiet = T)

ndvi <- rast("prac/MODIS/20220130_MODIS_v1g_POT_MOD13Q1_006_NDVI.tif")

# Get veg map and fire and plot
veg <- st_read("/Users/jasper/Documents/Datasets/SANBI/RLE_2021_Remnants/FynRemnantsClean.gpkg", quiet = T)
ndvi_g <- project(ndvi, st_crs(veg)$Wkt)
veg <- st_crop(veg, ndvi_g, quiet = T)
#fire <- st_read("/home/jasper/Documents/Datasets/Fire/All_Fires_20_21_gw/CNCFires2021Clean.gpkg", quiet = T)
#fire <- st_crop(fire, ndvi_g, quiet = T)
mapview::mapview(veg, zcol = "name", quiet = T) #+
#mapview::mapview(fire, zcol = "year", quiet = T)

```

---

## What are you going to measure?

### You decided that at each location you'd set up a 4x4m plot and...

Sample _Environmental conditions:_

- Estimate % projected cover (the area you would see from above) bare soil

- Take a densiometer reading at ground level

- Take a **soil sample** from the 4 corners to measure volumetric soil water content

- Note prevalence of **dung**: none/low/medium/high

- Note presence/absence of burnt dead skeletons or charcoal

- Note prevalence of serotiny
  
- Take a few notes (and photos) on any features you think may be interesting

---

## What are you going to measure?

### You decided that at each location you'd set up a 4x4m plot and...

Sample _Diversity:_

- Count the number of individuals and estimate % cover of each species

- Collect a voucher specimen of each species

- Assign species to functional types:
  
  - Trees, shrubs, herbs, grasses, succulents, geophytes, climbers, parasites, etc.?

---

class: center, middle

R code for the following analyses can be accessed [**at this link**](https://github.com/jslingsby/BIO3018F/blob/main/PracPresentation2025.Rmd)


---

### Some results from the survey:

```{r echo = F, fig.align = 'center', out.height = '60%', warning=F, messages=F}
dat = read_xlsx("prac/2025/Data_Biodiversity2_withdensiometer.xlsx", sheet = "Sheet1") # , range = "A1:M13")

dat %>%
  kable("html") %>% 
  kable_styling("striped", full_width = F)
```

---

### Some results from the survey 

```{r echo = F, fig.align = 'center', out.height = '60%', warning=F, messages=F}

set.seed(1234) #So I get the same results each time

vd = read_xlsx("prac/2025/Data_Biodiversity2_withdensiometer.xlsx", sheet = "Sheet2", range = "A1:DK14")

# vd = vd %>% select(Site, WorkingName) %>% 
#   distinct() %>% 
#   mutate(Abundance = abs(round(rnorm(nrow(.), 5, 2)))) 

vd %>%
  kable("html") %>% 
  kable_styling("striped", full_width = F)
 
```

---

### Get _count_ of species by site

```{r echo = F, fig.align = 'center', out.height = '60%', warning=F, messages=F}
# Do a pivot_longer to get species and abundance in separate columns and count species as records with abundance > 0 and present as a table
sn = vd %>% slice(-1) %>% # Drop the functional type row
  pivot_longer(names_to = "Species", cols = 2:115, values_to = "Abundance")  %>% # Convert to long format data
  group_by(Site) %>% # Set is so that summaries are grouped by Site
  summarize(`Species Number` = sum(Abundance > 0, na.rm=TRUE)) # Count the number of species (i.e. cells that are > 0)

sn %>%
  kable("html") %>% 
  kable_styling("striped", full_width = F)

```

---

### Add Shannon's Diversity Index (H) by site 

```{r echo = F, fig.align = 'center', out.height = '60%', warning=F, messages=F}

sdiv = vd %>% slice(-1) %>% # remove functinoal type
  select(-Site) %>% # remove site
  mutate_if(is.character, as.numeric) %>% # convert all data from class character to class numeric
  replace(is.na(.), 0) %>% # replace "NA" values with 0
  vegan::diversity(index="shannon") # calculate shannon's diversity

sn %>% 
  mutate(`Shannon Diversity` = sdiv) %>% # add shannon diversity to the species richness table
  kable("html") %>% 
  kable_styling("striped", full_width = F)

```

---

### Add Pielou's evenness (J) by site 

```{r echo = F, fig.align = 'center', out.height = '60%', warning=F, messages=F}

sn %>% 
  mutate(`Shannon Diversity` = sdiv) %>% # add shannon diversity to the species richness table
  mutate(`Pielou Evenness` = `Shannon Diversity`/log(`Species Number`)) %>% # calculate Pielou's evenness
  kable("html") %>% 
  kable_styling("striped", full_width = F)

```

---

### What about the number of individuals per site?

.pull-left[
```{r echo = F, fig.align = 'center', out.height = '100%', warning=F, messages=F}
sn_ni = vd %>% slice(-1) %>% # Drop the growth form row
  pivot_longer(names_to = "Species", cols = 2:115, values_to = "Abundance") %>% # Convert to long format data
  mutate_at("Abundance", as.numeric) %>% # Convert abundance data to numeric
  group_by(Site) %>% # Group by site
  summarize(`Species Number` = sum(Abundance > 0, na.rm=TRUE), 
            `Number of Individuals` = sum(Abundance, na.rm=TRUE)) # Count the number of species and number of individuals

sn_ni %>% 
  mutate(`Shannon Diversity` = sdiv) %>% # add shannon diversity to the species richness table
  mutate(`Pielou Evenness` = `Shannon Diversity`/log(`Species Number`)) %>% # calculate Pielou's evenness
  kable("html") %>% 
  kable_styling("striped", full_width = F)
```
]

.pull-right[

```{r echo = F, fig.align = 'center', out.height = '100%', warning=F, messages=F}
sn_ni %>% 
  ggplot(aes(y = `Species Number`, x = `Number of Individuals`)) + # Plot species number against number of individuals
  geom_text(label = sn_ni$Site) # Add site labels
```
]

---

### Let's add _rarefied_ species richness per site?

```{r echo = F, fig.align = 'center', out.height = '100%', warning=F, messages=F}
sr = vd %>% slice(-1) %>% # remove functinoal type
  select(-Site) %>% # remove site
  mutate_if(is.character, as.integer) %>% # convert all data from class character to class integer
  replace(is.na(.), 0) %>% # replace "NA" values with 0
  vegan::rarefy(10) # rarefy to 10 individuals (less than minimum number of individuals in any site)

sn_ni %>% 
  mutate(`Species Richness` = sr) %>% # add rarefied species richness to the table
  mutate(`Shannon Diversity` = sdiv) %>% # add shannon diversity to the species richness table
  mutate(`Pielou Evenness` = `Shannon Diversity`/log(`Species Number`)) %>% # calculate Pielou's evenness
  kable("html") %>% 
  kable_styling("striped", full_width = F)
```

---

### And compare the relationships among them...

```{r echo = F, fig.align = 'center', out.height = '100%', warning=F, messages=F}
sn_ni %>% 
  mutate(`Species Richness` = sr) %>% # add rarefied species richness to the table
  mutate(`Shannon Diversity` = sdiv) %>% # add shannon diversity to the species richness table
  mutate(`Pielou Evenness` = `Shannon Diversity`/log(`Species Number`)) %>% # calculate Pielou's evenness
  select(-Site) %>% # Drop Site
  ggpairs()
```

---

### What about compositional dissimilarity among sites? (presence/absence)

```{r echo = F, fig.align = 'center', out.height = '100%', warning=F, messages=F}

# Make new object omitting the functional types, make site names the row names and remove the site column
tvd = vd %>% slice(-1)
rownames(tvd) = tvd$Site

# Calculate Sorenson's distances between sites
sdist = tvd %>% select(-Site) %>% # remove site
  mutate_if(is.character, as.numeric) %>% # convert all data from class character to class numeric
  replace(is.na(.), 0) %>% # replace "NA" values with 0
  #as.matrix() %>% 
  designdist(method = "(A+B-2*J)/(A+B)", terms = "binary")

# some tomfoolery to present it nicely... (kable() doesn't like class "dist" - i.e. distance objects)
msite_BD <- as.data.frame(as.matrix(sdist)) # make it a rectangular table
msite_BD[upper.tri(msite_BD, diag = T)] <- NA # set the top-right to "NA"
options(knitr.kable.NA = '') # set the kable() table printing function to make NA values blank
rownames(msite_BD) <- unique(tvd$Site) # add names
colnames(msite_BD) <- rownames(msite_BD) # make the rownames the column names

# Present as a table
kable(msite_BD, digits = 3)
```

---

### What about compositional dissimilarity among sites? (abundance)

```{r echo = F, fig.align = 'center', out.height = '100%', warning=F, messages=F}

# Calculate Bray-Curtis distances between sites
sdist = tvd %>% select(-Site) %>% # remove site
  mutate_if(is.character, as.numeric) %>% # convert all data from class character to class numeric
  replace(is.na(.), 0) %>% # replace "NA" values with 0
  #as.matrix() %>% 
  designdist(method = "(A+B-2*J)/(A+B)", terms = "quadratic")

# some tomfoolery to present it nicely... (kable() doesn't like class "dist" - i.e. distance objects)
msite_BD <- as.data.frame(as.matrix(sdist)) # make it a rectangular table
msite_BD[upper.tri(msite_BD, diag = T)] <- NA # set the top-right to "NA"
options(knitr.kable.NA = '') # set the kable() table printing function to make NA values blank
rownames(msite_BD) <- unique(tvd$Site) # add names
colnames(msite_BD) <- rownames(msite_BD) # make the rownames the column names

# Present as a table
kable(msite_BD, digits = 3)
```

---

## References

```{r refs, echo=FALSE, results="asis"}
NoCite(myBib)
PrintBibliography(myBib)
```

---
class: center, middle

# Thanks!

Slides created via the R packages:

[**xaringan**](https://github.com/yihui/xaringan)<br>
[gadenbuie/xaringanthemer](https://github.com/gadenbuie/xaringanthemer)

The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](http://yihui.name/knitr), and [R Markdown](https://rmarkdown.rstudio.com).
